% R bootcamp, Module 3: Working with objects and data
% August 2018, UC Berkeley
% Chris Paciorek

```{r chunksetup, include=FALSE} 
# include any code here you don't want to show up in the document,
# e.g. package and dataset loading
if(!('modules' %in% unlist(strsplit(getwd(), split = '/')))) setwd('modules')
air <- read.csv(file.path('..', 'data', 'airline.csv'), stringsAsFactors = FALSE)
```

# Matrices and Arrays

Let's review matrices

- Matrices are two-dimensional collections of values of the same type
- We can have numeric, integer, character, or logical matrices, character matrices.
- You can't mix types within a matrix
```{r}
mat <- matrix(rnorm(12), nrow = 3, ncol = 4)
mat

# vectorized calcs work with matrices too
mat*4
mat <- cbind(mat, 1:3)
mat
```

Arrays are like matrices but can have more or fewer than two dimensions.
```{r}
arr <- array(rnorm(12), c(2, 3, 4))
arr
```

# Attributes

Objects have *attributes*.

```{r attr}
attributes(mat)
rownames(mat) <- c('first', 'middle', 'last')
mat
attributes(mat)

names(attributes(air))
attributes(air)$names
attributes(air)$row.names[1:10]
```

Now let's do a bit of manipulation and see if you can infer how R represents matrices internally.

```{r what_is_a_matrix}
mat[4]
attributes(mat) <- NULL
mat
is.matrix(mat)
```

**Question**: What can you infer about what a matrix is in R?

**Question**: What kind of object are the attributes themselves? How do I check?

# Matrices are stored column-major

This is like Fortran, MATLAB and Julia but not like C or Python(numpy). 

```{r}
mat <- matrix(1:12, 3, 4)
mat
c(mat)
```
You can go smoothly back and forth between a matrix (or an array) and a vector:
```{r}
identical(mat, matrix(c(mat), 3, 4))
identical(mat, matrix(c(mat), 3, 4, byrow = TRUE))
```

This is a common cause of bugs!


# Missing values and other special values

Since it was designed by statisticians, R handles missing values very well relative to other languages.

* `NA` is a missing value
```{r, fig.cap = ""}
vec <- rnorm(12)
vec[c(3, 5)] <- NA
vec
length(vec)
sum(vec)
sum(vec, na.rm = TRUE)
hist(vec)
is.na(vec)
```
Be careful because many R functions won't warn you that they are ignoring the missing values.

* To infinity and beyond
```{r}
big <- 1e500 
big
big + 7
```

* `NaN` stands for Not a Number
```{r}
sqrt(-5)
big - big
1/0
```

* `NULL`
```{r nulldata}
vec <- c(vec, NULL) 
vec
length(vec)
a <- NULL
a + 7
a[3, 4]
is.null(a)
myList <- list(a = 7, b = 5)
myList$a <- NULL  # works for data frames too
myList
```

`NA` can hold a place but `NULL` cannot.
`NULL` is useful for having a function argument default to 'nothing'. See `help(crossprod)`, which can compute either $X^{\top}X$ or $X^{\top}Y$.  


# Logical vectors
```{r}
answers <- c(TRUE, TRUE, FALSE, FALSE)
update <- c(TRUE, FALSE, TRUE, FALSE)

answers & update
answers | update
# note the vectorized boolean arithmetic

# what am I doing here?
sum(answers)
mean(answers)
answers + update
```

**Question**: What do you think R is doing to do arithmetic on logical vectors?

Tricks with logicals...

```{r}
identical(answers & update, as.logical(answers * update))
identical(answers | update, as.logical(answers + update))
```

# Data frames

A review from Module 1...

- Data frames are combinations of vectors of the same length, but can be of different types
- Data frames are what is used for standard rectangular (record by field) datasets, similar to a spreadsheet
- Data frames are a functionality that both sets R aside from some languages (e.g., Matlab) and provides functionality similar to some statistical packages (e.g., Stata, SAS)

```{r}
air <- read.csv('../data/airline.csv')
class(air)
head(air)
str(air)
```

# Data frames are (special) lists!

```{r}
is.list(air)
length(air)
air[[3]][1:5]
lapply(air, class) 
```

`lapply()` is a function used on lists; it works here to apply the `class()` function to each element of the list, which in this case is each field/column.

# But lists are also vectors!

```{r}
length(air)
someFields <- air[c(3,5)]
head(someFields)
identical(air[c(3,5)], air[ , c(3,5)])
```

In general the placement of commas in R is crucial, but here, two different operations give the same result because of the underlying structure of data frames.

# Factors
- A factor is a special data type in R used for categorical data. In some cases it works like magic and in others it is incredibly frustrating. 

```{r fac}
class(air$Dest)
head(air$Dest) # What order are the factors in?
levels(air[["Dest"]])  # note alternate way to get the variable
summary(air$Dest)
```

- What if we don't like the order these are in? Factor order is important for all kinds of things like plotting, analysis of variance, regression output, and more

# Ordering the Factor
- Ordered factors simply have an additional attribute explaining the order of the levels of a factor
- This is a useful shortcut when we want to preserve some of the meaning provided by the order
- Think ordinal data

```{r orderedfac, fig.cap = ""}
# first let's do some pre-processing to get set up for the example
air$DepDelayCens <- air$DepDelay
air$DepDelayCens[air$DepDelayCens > 60] <- 60
air$DepDelayCens[air$DepDelayCens < 0] <- 0
air$DayOfWeek <- as.factor(air$DayOfWeek)
air$DayOfWeekSun <- ordered(air$DayOfWeek, 
     levels = levels(air$DayOfWeek)[c(7,1,2,3,4,5,6)])

# alternative coding
# air <- within(air, 
#     DayOfWeekSun <- ordered(DayOfWeek, 
#        levels = levels(DayOfWeek)[c(7,1,2,3,4,5,6)])
#)

head(air$DayOfWeekSun)
levels(air$DayOfWeekSun)
levels(air$DayOfWeekSun) <- c("Sun","Mon","Tue","Wed","Thu","Fri","Sat")
boxplot(DepDelayCens ~ DayOfWeekSun, data = air)
```

**Challenge**: Try to decipher what I just did with that complicated single line of code starting with `within`.

# Reclassifying Factors
- Turning factors into other data types can be tricky. All factors have an underlying numeric structure.

```{r fac2}
students <- factor(c('basic','proficient','advanced','basic', 
      'advanced', 'minimal'))
levels(students)
unclass(students)
```

- Hmmm, what happened?
- Be careful! The best way to convert a factor is to convert it to a character first.

```{r}
students <- factor(c('basic','proficient','advanced','basic', 
      'advanced', 'minimal'))
score = c(minimal = 3, basic = 1, advanced = 13, proficient = 7) # a named vector
score["advanced"]  # look up by name
students[3]
score[students[3]]
score[as.character(students[3])]
```

What went wrong and how did we fix it?  Notice how easily this could be a big bug in your code.

# Subsetting

There are many ways to select subsets in R. The syntax below is useful for vectors, matrices, data frames, arrays and lists.

```{r}
vec <- rnorm(20)
mat <- matrix(vec, 4, 5)
rownames(mat) <- letters[1:4]
mat
```
1) by direct indexing

```{r}
vec[c(3, 5, 12:14)]
vec[-c(3,5)]
mat[c(2,4), 5]
rowInd <- c(1, 3, 4)
colInd <- c(2, 2, 1)
elemInd <- cbind(rowInd, colInd)
elemInd
mat[elemInd]
```

Note the last usage where we give it a 2-column matrix of indices

2) by a vector of logicals

```{r}
cond <- vec > 0
vec[cond]
air[air$DepDelay > 60*12 & !is.na(air$DepDelay), ]
```

What happened in the last subsetting operation?

3) by a vector of names
```{r}
mat[c('a', 'd', 'a'), ]
```
4) using *subset()*

```{r}
subset(air, DepDelay > 60*12)
```

5) using *dplyr* tools such as *filter()* and *select()* -- more in Module 5

# Assignment into subsets

We can assign into subsets by using similar syntax, as we saw with vectors.

```{r}
vec[c(3, 5, 12:14)] <- 1:5
mat[2, 3:5] <- rnorm(3)
mat[mat[,1] > 0, ] <- -Inf
```

# Strings

R has lots of functionality for character strings. Usually these are stored as vectors of strings, each with arbitrary length.

```{r}
chars <- c('hi', 'hallo', "mother's", 'father\'s', "He said, \"hi\"" )
length(chars)
nchar(chars)
paste("bill", "clinton", sep = " ")  # paste together a set of strings
paste(chars, collapse = ' ')  # paste together things from a vector

strsplit("This is the R bootcamp", split = " ")
substring(chars, 2, 3)
chars2 <- chars
substring(chars2, 2, 3) <- "ZZ"
chars2
```
We can search for patterns in character vectors and replace patterns (both vectorized!)
```{r}
grep("ther", chars)
gsub("hi", "Hi", chars)
```

# Regular expressions (regex or regexp)

Some of you may be familiar with using *regular expressions*, which is functionality for doing sophisticated pattern matching and replacement with strings. *Python* and *Perl* are both used extensively for such text manipulation. 

R has a full set of regular expression capabilities available through the *grep()*, *gregexpr()*, and *gsub()* functions (among others - many R functions will work with regular expressions). A particularly nice way to make use of this functionality is to use the *stringr* package, which is more user-friendly than directly using the core R functions.

You can basically do any regular expression/string manipulations in R, though the syntax may be a bit clunky at times.

# More details on reading data into R

Remember that you'll need to know the current working directory so that you know where R is looking for files.

The workhorse for reading into a data frame is *read.table()*, which allows any separator (CSV, tab-delimited, etc.). *read.csv()* is a special case of *read.table()* for CSV files.

You've already seen a bit of this, but let's work through a more involved example, so you can see some of the steps and tricks involved in reading data into R.

```{r, cache=TRUE}
rta <- read.table("../data/RTAData.csv", sep = ",", head = TRUE)
rta[1:5, 1:5]
dim(rta)
# great, we're all set, right?
# Not so fast...
unlist(lapply(rta, class))[1:5]
# ?read.table
rta2 <- read.table("../data/RTAData.csv", sep = ",", 
  head = TRUE, stringsAsFactors = FALSE)
rta2[3,3]
unlist(lapply(rta2, class))[1:5]
# let's delve more deeply
levels(rta[ , 2])[c(1:5, 3041:3044)]
rta3 <- read.table("../data/RTAData.csv", sep = ",", head = TRUE, 
      stringsAsFactors = FALSE, na.strings = c('NA', 'x'))
unlist(lapply(rta3, class))[1:5]

# checking...
missing <- which(rta[ , 2] == "")
missing[1:5]
rta3[head(missing), ]
```

It's good to first look at your data in plain text format outside of R and then to check it after you've read it into R.

# Other ways to read data into R

The *read.table()* family of functions just skims the surface of things...

1) You can also read in a file as vector of characters, one character string per line of the file with *readLines()*, and then post-process it. 
2) You can read fixed width format (constant number of characters per field) with *read.fwf()*.
3) *read_csv()* (and *read_lines()*, *read_fwf()*, etc.) in the *readr* package is a faster, more helpful drop-in replacement for *read.csv()* that plays well with *dplyr* (see Module 5).
4) the *data.table* package is great for reading and manipulating large datasets (orders of gigabytes or 10s of gigabytes).


# Breakout

### Basics

1) Extract the 5th row from the airline dataset.

2) Extract the last row from the airline dataset.

3) Count the number of NAs in the departure delay field of the airline dataset.

4) Set all of the extreme delays (more than 300 minutes) to NA.

5) Consider the first row of the airline dataset, which has UA flight number 548. How do I create a string "UA-548" using `air$UniqueCarrier[1]` and `air$FlightNum[1]`? 

### Using the ideas

6) Create a character string using `paste()` that tells the user how many rows there are in the data frame - do this programmatically such that it would work for any data frame regardless of how many rows it has. The result should look like this: "There are 56,234,234 rows in the dataset"

7) If you didn't do it this way already, extract the last row from the airline dataset without typing the number '539895'.

8) Create a boolean vector indicating if flight on the carrier WN (Southwest) and distance is less than 1000 miles and calculate the proportion of all the flights these represent.

9) Use that vector to create a new data frame that is a subset of the original data frame.

10) Consider the attributes of the *air* dataset. What kind of R object is the set of attributes?

### Advanced

11) Make the destination variable a factor variable and reorder the levels by the number of arrivals from SFO at the destinations. Now create a boxplot of departure delay by destination for January, 2005.

12) Create row names for the data frame based on concatenating the Year, Month, DayOfMonth, UniqueCarrier and FlightNum fields. You may run into a problem with the data that you need to resolve before you're able to do this...
